---
title: "Comparative Modeling Approaches for Breast Cancer Prognosis --- From Gene Selection with LASSO to PCA and VAE-Based Dimensionality Reduction"
author: "Baodong Zhang"
editor: visual
format:
  pdf:
    documentclass: article
    fontsize: 6pt
echo: true
warning: false
message: false
fig-width: 6
fig-height: 4
fig-align: center
geometry: "top=1in, bottom=1in, left=1in, right=1in"
header-includes:
  - \usepackage[table,xcdraw]{xcolor}
  
bibliography: references.bib
---

```{r global-setup, include=FALSE}

Sys.setenv(
  OMP_NUM_THREADS = "1",
  MKL_NUM_THREADS = "1",
  OPENBLAS_NUM_THREADS = "1",
  VECLIB_MAXIMUM_THREADS = "1",
  NUMEXPR_NUM_THREADS = "1"
)
# if still libiomp conflict：
# Sys.setenv(KMP_DUPLICATE_LIB_OK = "TRUE")

knitr::opts_chunk$set(
  echo = FALSE,       
  eval = TRUE,       
  warning = FALSE,   
  message = FALSE,   
  error = FALSE,     
  results = 'markup' 
  )
library(dplyr)
library(survival)
library(survminer)
library(glmnet)
library(tibble)
library(kableExtra)
library(mice)
library(randomForestSRC)
library(riskRegression) 
library(caret)
library(pec) 

if (requireNamespace("data.table", quietly = TRUE)) data.table::setDTthreads(1)
if (requireNamespace("RhpcBLASctl", quietly = TRUE)) {
  RhpcBLASctl::omp_set_num_threads(1)
  RhpcBLASctl::blas_set_num_threads(1)
}

source("qmd_functions.R")
#source("Project_functions.R")
#options(scipen = 999)

```

## Summarize

This project evaluated multiple approaches to predicting distant recurrence-free survival (DRFS) in breast cancer using genomic data. Starting with more than 20,000 genes, univariate Cox regression identified several thousand candidates, which were further reduced to a few dozen genes using LASSO-Cox. We compared model performance across three input sets: (i) clinical variables only; (ii) clinical variables plus gene-expression data; and (iii) gene-expression data only, using both Cox proportional hazards models and Random Survival Forests (RSF). We also tested dimensionality-reduction methods—principal component analysis (PCA) and a variational autoencoder (VAE)—applied to the gene-expression data; the resulting principal components and VAE latent variables were subsequently modeled with Cox regression.

Extensive bootstrap cross-validation was used to evaluate models by mean time-dependent AUC (iAUC) and concordance index (C-index) across iterations. Data preprocessing included Multivariate Imputation by Chained Equations (MICE) for missing clinical values, z-score normalization of gene-expression data, and strict separation of training and validation sets to prevent data leakage. Across 100 bootstrap resamples with replacement, none of the models achieved strong performance on the test set. By comparison, the PCA-based Cox model modestly outperformed the others (mean iAUC = 0.57; mean C-index = 0.56). The single best PCA–Cox model—identified on one train/test split by the highest combined score-was retained as the final model for future prediction.

## Introduction

Breast cancer remains a leading threat to women’s health. Traditional survival analysis in breast cancer has relied primarily on clinicopathologic variables such as tumor stage, histologic grade, lymph-node status, and receptor status. However, even the strongest predictors of metastasis—for example, nodal status and histologic grade—often fail to accurately stratify tumors by their clinical behavior.[@vantVeer2002GeneExpressionProfiling] As high-throughput sequencing has become widespread and more affordable, genomic data have become central to predicting outcomes and guiding treatment decisions. The tumor’s molecular profile—specific mutations and gene-expression patterns—drives its behavior, which in turn leads to different clinical trajectories. [@Venetis2024EvolvingGenomicTesting]

Transcriptomic datasets commonly used in survival studies include RNA-seq and microarrays. Standard methods in survival analysis include Kaplan–Meier curves with the log-rank test and multivariable Cox proportional hazards models. Regularized Cox models (e.g., LASSO-Cox) and Random Survival Forests are also widely applied, including for feature selection. Researchers use unsupervised clustering to identify molecular subtypes associated with distinct outcomes. In recent years, deep learning–based survival methods have gained popularity for multimodal data, complementing traditional statistical models.[@Wiegrebe2024DeepLearningSurvival]

In this project, we built and compared classical survival models—specifically Cox proportional hazards models and Random Survival Forests—and assessed the predictive contribution of genomic features relative to clinical variables. For genomic data, beyond selecting predictive genes via feature engineering for direct modeling, we applied PCA for dimensionality reduction and trained a VAE to learn compact latent representations that aew subsequently entered into Cox models. During exploratory analyses, patient samples were clustered by gene expression to define molecular subtypes, and survival differences across subtypes were evaluated.

## Data Source

The data for this project were sourced from the internationally renowned NCBI GEO Database (Gene Expression Omnibus) under the accession number GSE7390, which includes a total of 198 breast cancer patients. All patients were lymph node-negative. Clinical variables include age, tumor grade, estrogen receptor status (ER status),and tumor size. The gene expression data were obtained using microarray technology and comprise 22,283 variables. The target variables include: Distant Metastasis-Free Survival Status(e.dmfs): A binary variable indicating whether the patient experienced disease recurrence (yes/no). Time to Distant Metastasis-Free Survival(t.dmfs): The duration from treatment to the occurrence of disease recurrence, measured in days. For patients who did not experience recurrence, the last known recurrence-free time was recorded (i.e., right-censored data).

```{r}

expr <- readRDS("expr_7390.RDS")
clinical <- readRDS("clinical_7390.rds")
subset_tibble <- as_tibble(expr[1:5, 1:5], rownames = "Probe_ID")

# process clinical data
clinical$time <- as.numeric(clinical$`t.dmfs:ch1`)
clinical$status <- as.numeric(clinical$`e.dmfs:ch1`)

clinical_selected <- clinical[, c(
  "geo_accession",
  "t.dmfs:ch1",
  "e.dmfs:ch1",
  "age:ch1",
  "er:ch1",
  "grade:ch1",
  "size:ch1",
  "node:ch1",
  "hospital:ch1", "samplename:ch1", "filename:ch1"
)]

original_names <- colnames(clinical_selected)
clean_names <- gsub(":ch1", "", original_names)
clean_names <- gsub("\\.", "_", clean_names)
colnames(clinical_selected) <- clean_names

clinical_cleaned <- as.data.frame(lapply(clinical_selected, function(x) {
  x_num <- suppressWarnings(as.numeric(as.character(x)))
  if (all(is.na(x_num))) x else x_num
}))

clinical_cleaned_7390 <- clinical_cleaned %>%
  select(geo_accession, t_dmfs, e_dmfs, age, er, grade, size, hospital)


kable_table_func(subset_tibble, "Subset of Gene Expression Matrix (n = 5)")
kable_table_func(head(clinical_cleaned_7390, 5), "Subset of Clinical Data (n = 5)")

```

## Data Preprocessing and Exploratory Data Analysis (EDA)

### Clean clinical data: handle missing values and outliers

```{r}

numeric_vars <- names(clinical_cleaned_7390)[sapply(clinical_cleaned_7390, is.numeric)]
generate_box_plots(clinical_cleaned_7390, numeric_vars[6])

```

The box plot of clinical data shows no outliers.

```{r}
# Detect and define factor variables appropriately:
for (i in 4:7) {
 if (length(unique(clinical_cleaned_7390[, i])) <= 4) {
   clinical_cleaned_7390[, i] <- factor(clinical_cleaned_7390[, i])
 }
}
# Detect missed values
na_info <- colSums(is.na(clinical_cleaned_7390))
na_info <- na_info[na_info > 0]
na_df <- data.frame(
  Column = names(na_info),
  NA_Count = as.numeric(na_info),
  row.names = NULL
)
na_df$Class <- "factors level > 2"
na_df$Mice_Method <- "polyreg"
kable_table_func(na_df, "Table 3: NA Audit Report")
```

The clinical dataset exhibited missingness exclusively in the grade column, with two incomplete observations. As grade represents an ordinal factor variable with three levels, multiple imputation was performed using the proportional odds logistic regression ("polr") method within the MICE framework. The final imputed values were derived through majority voting across the multiply imputed datasets.To prevent data leakage, the MICE algorithm was trained solely on the training subset. The validation set was subsequently imputed using the parameters estimated from the training data, thereby preserving the validation set's integrity for unbiased model performance assessment.

```{r include=FALSE}
# miss data imputation
clinical_cleaned_7390_imputed <- impute_missing_value(clinical_cleaned_7390, selected_col= c("age", "er", "grade", "size"), missed_col = "grade")
before_imp <- clinical_cleaned_7390 %>% filter(geo_accession %in% c("GSM178004", "GSM178011")) %>% mutate(status = "before imputation")
after_imp <- clinical_cleaned_7390_imputed %>% filter(geo_accession %in% c("GSM178004", "GSM178011")) %>% mutate(status = "after imputation")
na_imp_result <- rbind(before_imp, after_imp)
kable_table_func(na_imp_result, "Missing Value Imputation with MICE")
hospital_counts <- table(clinical_cleaned_7390_imputed$hospital)
```

There are `r length(unique(clinical_cleaned_7390$hospital))` hospitals in the clinical dataset. The Kaplan-Meier survival curves generated by ggsurvplot demonstrate partially overlapping patterns across these hospitals, and the log-rank test yields a p-value of 0.1. While this suggests some degree of heterogeneity in survival outcomes among different hospital sites, the p-value \> 0.05 indicates that the observed differences are not statistically significant at the conventional alpha level of 0.05. Given the limited sample size and the lack of statistical significance, the random effects associated with hospital variation will not be considered as a primary focus in subsequent analyses.

```{r}
# Kaplan-Meier log-rank test on "hospital random effect"
fit <- survfit(Surv(t_dmfs, e_dmfs) ~ hospital, data = clinical_cleaned_7390_imputed)
ggsurvplot(fit, data = clinical_cleaned_7390_imputed, pval = TRUE) 
```

### Exploring the gene data with hierarchical cluster

The value of gene expression ranges from `r min(expr)` to `r max(expr)` and has multiple decimal places. This range is typical data converted by Log2. The summary table below indicates the absence of both outliers and missing values.

```{r}
row_means <- rowMeans(expr, na.rm = TRUE)
row_sds <- apply(expr, 1, sd, na.rm = TRUE)
summary_df <- rbind(
  data.frame(Statistic = "Mean", as.list(summary(row_means))),
  data.frame(Statistic = "SD", as.list(summary(row_sds)))
)

kable_table_func(summary_df, "Summary of Genomic Data")

```

```{r}
# Data quality check
different_samples <- setdiff(colnames(expr), clinical_cleaned_7390$geo_accession)
if (!length(different_samples) == 0){
  print("The expression matrix and clinical data DOES NOT have matching sample IDs")
}

# Given computational constraints select top 1500 most variable genes 
gene_vars <- apply(expr, 1, var) 
n_gene <- 1500
top_gene_index <- order(gene_vars, decreasing = TRUE)[1:n_gene] 
expr_filtered <- expr[top_gene_index, ] 
```

Leveraging genomic features allows us to stratify patients into distinct molecular subgroups. We then assess whether survival curves differ meaningfully across these clusters, aiding the identification of molecular determinants that influence prognosis.

Because of computational constraints, we first performed feature preselection by taking the top `r n_gene` most variable genes (variance ranking). We computed pairwise Euclidean distances to form a dissimilarity matrix and applied hierarchical clustering using Ward’s minimum-variance linkage, which promotes compact, low–within-cluster variance groupings.

```{r}
#Calculate distance matrix between samples(patients) features are genes
sample_dist <- dist(t(expr_filtered), method = "euclidean") 
# Perform hierarchical clustering using Ward's method which minimizes within-cluster variance
sample_hclust <- hclust(sample_dist, method = "ward.D2")
plot(sample_hclust, 
     labels = FALSE,          # Hide sample labels for clarity
     main = "Sample Hierarchical Clustering Dendrogram",
     xlab = "Samples", 
     ylab = "Height (Distance)")
```

To determine the optimal clustering resolution, cluster numbers ranging from 2 to 31 were systematically evaluated. For each clustering solution, Kaplan-Meier survival analysis was conducted accompanied by log-rank testing to assess inter-cluster survival differences.

```{r}
p_values <- numeric(30)  # Store p-values for k=2 to 50
best_k <- NULL
best_p_value <- Inf

# Iterate over k from 2 to 50
for (k in 2:31) {
  # Cut dendrogram to define k molecular subtypes
  clusters <- cutree(sample_hclust, k = k)
  
  # Add cluster information to clinical data
  clinical_cleaned_7390$Molecular_Subtype <- factor(clusters, 
                                                   labels = paste0("Subtype_", 1:k))
  
  # Create survival object
  surv_obj <- Surv(time = clinical_cleaned_7390$t_dmfs, 
                  event = clinical_cleaned_7390$e_dmfs)
  
  # Fit survival curves
  fit <- survfit(surv_obj ~ Molecular_Subtype, data = clinical_cleaned_7390)
  
  # Perform log-rank test and extract p-value
  logrank_test <- survdiff(surv_obj ~ Molecular_Subtype, data = clinical_cleaned_7390)
  p_value <- 1 - pchisq(logrank_test$chisq, df = length(logrank_test$n) - 1)
  
  # Store p-value
  p_values[k-1] <- p_value
  
  # Update best k if current p-value is lower
  if (p_value < best_p_value) {
    best_p_value <- p_value
    best_k <- k
  }
  # Print progress
  # cat(sprintf("k = %d: p-value = %.4f\n", k, p_value))
}

# cat(sprintf("\nOptimal k: %d with minimum p-value: %.6f\n", best_k, best_p_value))

# Create results dataframe
results_df <- data.frame(
  k = 2:31,
  p_value = p_values
)

# Plot p-values vs k
ggplot(results_df, aes(x = k, y = -log10(p_value))) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = best_k, linetype = "dashed", color = "red") +
  annotate("text", x = best_k, y = max(-log10(p_values)), 
           label = paste("Optimal k =", best_k), color = "red") +
  labs(title = "P-value vs Number of Clusters (k)",
       x = "Number of Clusters (k)",
       y = "-log10(p-value)") +
  theme_minimal()

```

The optimal partition `r best_k` was selected based on the most statistically significant separation of survival curves, as determined by the minimal log-rank p-value across all tested configurations. Following the division of samples into `r best_k` groups, the KM log-rank test yielded a p-value of `r best_p_value`, indicating no statistically significant differences in survival outcomes among these groups.

```{r}

k = best_k
clusters <- cutree(sample_hclust, k = k)

# Add cluster information to clinical data as a new factor variable
clinical_cleaned_7390$Molecular_Subtype <- factor(clusters,
                                              labels = paste0("Subtype_", 1:k))

surv_obj <- Surv(time = clinical_cleaned_7390$t_dmfs,
                 event = clinical_cleaned_7390$e_dmfs)

# Fit survival curves stratified by molecular subtype
fit <- survfit(surv_obj ~ Molecular_Subtype, data = clinical_cleaned_7390)

# Generate Kaplan-Meier survival plot with statistical analysis
surv_plot <- ggsurvplot(fit,
           data = clinical_cleaned_7390,
           pval = TRUE,
          # pval.coord = c(50, 0.9),
           conf.int = FALSE,
           risk.table = FALSE,
           risk.table.height = 0.25,
           legend.labs = levels(clinical_cleaned_7390$Molecular_Subtype),
           legend.title = "Molecular Subtype",
           palette = "jama",
           xlab = "Time (Days)", 
           ylab = "Overall Survival Probability",
           title = "Survival Analysis by Molecular Subtype",
           ggtheme = theme_minimal())

print(surv_plot)
clinical_cleaned_7390 <- clinical_cleaned_7390 %>% select(-Molecular_Subtype) 
```

### Train Test Spliting and MAD filter on genomic data

Further processing will include: train test data splitting, clinical miss data imputation, genetic feature pre-selection, and standardization of genomic data.

```{r}
set.seed(92000028)
expr_mat <- expr
clinical_df <- clinical_cleaned_7390
# perform train test split and mice
data_prepare_result <- bootstrap_sampl_split_miss_vale_imputation(expr_mat, clinical_df)
train_expr <- data_prepare_result$train_expr
test_expr <- data_prepare_result$test_expr
train_clinical <- data_prepare_result$train_clinical
test_clinical <- data_prepare_result$test_clinical
events_train <- data_prepare_result$events_train

# quality check
test_clinical_unimputed = clinical_df %>% filter( geo_accession %in% test_clinical$geo_accession)
# setdiff(test_clinical_unimputed, test_clinical)
# setdiff(test_clinical, test_clinical_unimputed)
# MAD filter
quantile_cutoff = 0.25
filtered_data <- filter_genes_by_mad(train_expr, test_expr, quantile_cutoff)
train_expr2 <- filtered_data$train_expr
test_expr2 <- filtered_data$test_expr
```

Trough resampling with replacement, we performed train–test splits and applied multiple imputation to the clinical covariates. For the genomic features, we applied median absolute deviation (MAD) filtering on the training set to remove genes with minimal expression variability. Genes in the top `r 1-quantile_cutoff` quantile of variability were retained for downstream analysis. After preprocessing, the number of gene features was reduced from from `r nrow(expr_mat)` to `r nrow(train_expr2)`. The validation set was restricted to the same gene subset to ensure analytical consistency and prevent data leakage.

```{r}
#kable_table_func(train_clinical[1:5,], "Training set samples for clinical data (n = 5)")
#kable_table_func(train_expr2[1:5, 1:5], "Training set samples for gene expression data (n = 5)")
```

## COX Regression

### Univariate Cox regression

We performed univariate Cox regression to screen for prognostic genes. The cox proportional hazards model is a semi-parametric model that relates survival time to multiple predictors. Its specification is:

$$
  h(t|X) = h_0(t) \times \exp(\beta_1X_1 + \beta_2X_2 + \cdots + \beta_pX_p)
$$ where: - $h(t|X)$ is the hazard function at time $t$ given covariates $X$ - $h_0(t)$ is the baseline hazard function - $\exp(\beta_iX_i)$ represents the hazard ratio for each covariate

The univariate cox proportional hazards model just has one predictor: $$ h(t|X_1) = h_0(t) \times \exp(\beta_1X_1) $$

```{r}
    sig_gene_df <- batch_univariate_cox_regression(train_expr2, train_clinical, p_value = 0.05)
    message(sprintf("Univariate cox regression finished, significant genes = %d", if (is.null(sig_gene_df)) 0 else nrow(sig_gene_df)))

```

We fit the univariate Cox PH models for each gene. Genes showing a significant association (Wald z-test, 𝑝\< 0.05) were identified and compiled into a results table. The number of gene features was further reduced to `r nrow(sig_gene_df)`.

```{r}
sig_gene_df <- sig_gene_df %>%
  select(gene, everything()) %>%
  remove_rownames()

kable_table_func(head(sig_gene_df, 5), "UnivariateCox Significant Genes (n = top 5)")
```

### LASSO Cox PH regression

To further refine the feature set, we applied LASSO-penalized Cox PH regression (Least Absolute Shrinkage and Selection Operator). The L1 penalty enables simultaneous coefficient shrinkage and variable selection, yielding a parsimonious subset of prognostic genes, with the penalty strength tuned during model training. The objective function combines partial likelihood with L1 penalty:

$$
\min_{\beta} \left[ -\frac{1}{n} \ell(\beta) + \lambda \sum_{j=1}^{p} |\beta_j| \right]
$$

where:

-   $\ell(\beta)$ = partial likelihood of the Cox model
-   $\lambda$ = regularization parameter controlling shrinkage
-   $\sum_{j=1}^{p} |\beta_j|$ = L1 penalty term

For the multivariable or Lasso Cox survival modeling it is essential to compare effect sizes (hazard ratios) on a common scale. Therefore, Z-score normalization was applied individually to each gene and numerical clinical variable. Additional feature selection was performed using remove_high_corr_genes function with a correlation coefficient cutoff of 0.90 to address multicollinearity among predictors. To similarly prevent information leakage, the validation set was standardized utilizing the mean and standard deviation parameters calculated solely from the training dataset.

```{r}

# prepare data for Lasso Cox regression
significant_gene <- sig_gene_df$gene
train_expr_scaled <- standardize_with_train(gene_mat_train = train_expr2,
                                                gene_mat_valid = train_expr2,
                                                significant_gene = significant_gene)


train_expr_filtered <- remove_high_corr_genes(train_expr_scaled, cutoff = 0.90)
message("Correlation-based filtering completed: train_expr_filtered dimensions = ", 
        paste(dim(train_expr_filtered), collapse = " x "))

significant_gene2 <- rownames(train_expr_filtered)
message(sprintf("Successfully extracted filtered gene set, significant_gene2 contains %d features", 
                length(significant_gene2)))

# Normalize test set gene expression data using training set parameters
test_expr_scaled <- standardize_with_train(gene_mat_train = train_expr2,
                                           gene_mat_valid = test_expr2,
                                           significant_gene = significant_gene2)

kable_table_func((test_expr_scaled %>% 
   as.data.frame() %>% 
   rownames_to_column("gene"))[1:5, 1:6], "Genomic test set (n = 5), scaled (training-set parameters)")

# Normalized clinical data
columns_to_remove <- c("e_dmfs", "t_dmfs", "geo_accession", "hospital")

numeric_columns <- train_clinical %>%
  select(where(is.numeric), -any_of(columns_to_remove)) %>%
  names()

train_clinical_scaled <- standardize_with_train_clinical(train_clinical,
                                                             train_clinical,
                                                             scale_cols = numeric_columns)
test_clinical_scaled <- standardize_with_train_clinical(train_clinical,
                                                            test_clinical,
                                                            scale_cols = numeric_columns)
kable_table_func((test_clinical_scaled)[1:5,], "Clinical test set (n = 5), scaled (training-set parameters)")

```

```{r}
# Lasso Cox
gene_freq_df <- repeat_cv_lasso_cox(train_expr = train_expr_filtered,
                                        train_clinical,
                                        significant_gene_vec = significant_gene2,
                                        repeats = 5,
                                        nfolds = 10,
                                        alpha = 1)
message(sprintf("LASSO Cox finished"))
```

```{r}
min_epv <- 5 
min_freq <- 0.8
# EPV rule of thumb:
max_vars_allowed <- floor(events_train / min_epv) 
```

The LASSO Cox regression procedure incorporated repeated cross-validation to enhance the robustness of gene selection across multiple iterations. The resulting genes were prioritized based on selection frequency and mean coefficient magnitude, with a retention threshold set at ≥ `r min_freq` frequency. Regarding sample size considerations, Peduzzi et al. (1995, J Clin Epidemiol)[@Peduzzi1995EPVPH] initially proposed a conservative requirement of 10 events per variable (EPV) for Cox regression. Vittinghoff & McCulloch (2007, Am J Epidemiol) [@Vittinghoff2007RelaxingRule]later demonstrated that 5–9 EPV may be sufficient for many analytical scenarios. Consistent with the latter guidelines and considering the cohort size limitations, an EPV threshold of `r min_epv` was implemented for additional feature selection. Since there were `r events_train` events in the training data, the maximum number of variables allowed is `r max_vars_allowed` . Consequently, `r max_vars_allowed` genes were ultimately selected as feature variables.

However to assess whether adding gene-expression features improves the discriminative performance of Cox models over clinical covariates, we appended the genomic features directly to the clinical set without enforcing the events-per-variable (EPV) rule. The EPV constraint was likewise not re-applied after PCA-based dimensionality reduction in the next part. Owing to space limitations, a systematic evaluation of EPV-constrained specifications is left for future work.

```{r}
# Keep genes appearing at least 'min_freq' times across resamples
gene_freq_df_best <- gene_freq_df %>%
  filter(freq >= min_freq)          

# Take the top genes up to the EPV cap; attach their mean coefficients
selected_gene_df <- gene_freq_df[1:min(nrow(gene_freq_df_best), max_vars_allowed), ] 
kable_table_func(selected_gene_df, "Multi_Iteration_LASSO_Genes_Selection_Frequency_0.8_EPV_5")
selected_gene_df <- selected_gene_df %>%
  mutate(coef = mean_coef)
```

### Risk score calculation and Kplan Maier analysis between groups

The risk score for patient was calculated as:

$$
\text{risk score} = \beta_1 x_1 + \beta_2 x_2 + \dots + \beta_i x_i + \dots + \beta_p x_p
$$ where: - $\beta_i$ represents the coefficient for each variable - $x_i$ represents the value of each predictor variable - $p$ is the total number of variables in the model

Subsequently, patients were stratified into high-risk and low-risk groups based on the median risk score of the training cohort. This same median value, derived from the training set, was then applied as the cutoff to classify patients in the validation set into high- and low-risk groups following the calculation of their risk scores.

```{r}
result <- compute_risk_score_train_2(
      gene_mat_scaled  = train_expr_filtered,
      significant_vars_df = selected_gene_df,
      clinical_cleaned = train_clinical_scaled
      )

clinical_cleaned_risk_train <- result$merged_df
median_cut_train <- result$median_cut

clinical_cleaned_risk_test <- compute_risk_score_test_2(
  gene_mat_scaled = test_expr_scaled,
  significant_vars_df = selected_gene_df,
  clinical_cleaned = test_clinical_scaled,
  median_cut_train)

clinical_cleaned_risk_test$group_var <- as.factor(clinical_cleaned_risk_test[["risk_group"]])
  
surv_obj <- Surv(time = clinical_cleaned_risk_test$t_dmfs, 
                   event = clinical_cleaned_risk_test$e_dmfs)
  
  
# Fit survival model
fit <- survfit(surv_obj ~ risk_group, data = clinical_cleaned_risk_test)
  
# Create survival plot
ggsurvplot(
  fit,
  data = clinical_cleaned_risk_test,
  pval = TRUE,
  conf.int = TRUE,
  risk.table = FALSE,
  xlab = "Days",
  ylab = "Survival Probability",
  palette = "Set1"
  )
  
surv_diff <- survdiff(surv_obj ~ group_var, data = clinical_cleaned_risk_test)
p_value <- 1 - pchisq(surv_diff$chisq, length(surv_diff$n) - 1)
# cat("Log-rank Kaplan-Meier test p-value:", p_value, "\n")

```

The Kaplan–Meier survival curves show substantial overlap between the two groups, and the log-rank test confirms no statistically significant difference in survival (p = `r p_value`).This indicates that stratifying patients by risk score does not yield a clear survival separation between groups.

### Safeguards Against Separation in Cox Models

In certain traning datasets, Cox models can exhibit (quasi-)complete separation, causing coefficients to diverge and the concordance index (C-index) to appear spuriously high. To screen out such pathological fits, we enforce two checks: if the model’s C-index exceeds min_concord (e.g., 0.97) or any coefficient’s absolute value exceeds coef_max (e.g., 10), the model is rejected. The following shows the model’s training results from one random train–test split. The validation of the model is presented following the section on Model Discrimination Assessment Metrics.

```{r}
# prepare data for cox modeling
columns_to_remove2 <- c("risk_group", "risk_score", "e_dmfs", "t_dmfs") 

character_columns <- clinical_cleaned_risk_train %>%
  select(where(is.character)) %>%
  colnames()

predictors <- colnames(clinical_cleaned_risk_train) %>%
  setdiff(c(columns_to_remove2, character_columns)) 

df <- clinical_cleaned_risk_train %>%
  select(c(predictors, "t_dmfs", "e_dmfs"))

df_test <- clinical_cleaned_risk_test %>%
  select(c(predictors, "t_dmfs", "e_dmfs"))

# fit cox
results_train <- fit_cox_model(predictors, df)

full_cox_result <- results_train$coef_table

kable_table_func(full_cox_result, "Full Cox Model Analysis Summary")

full_cox_ph_test <- as.data.frame(results_train$ph_test$table)

kable_table_func(full_cox_ph_test, "Full Cox Model Proportional Hazards Test")

ph_plots <- ggcoxzph(results_train$ph_test)
print(ph_plots[[1]])
# print(ph_plots[[2]])
# print(ph_plots[[3]])
# print(ph_plots[[4]])
print(ph_plots[[6]])

coef_max <- 10
min_concord <- 0.97


# detection of divergent coefficients in the Cox model
if (any(abs(coef(results_train$model)) > coef_max)) {
  message(sprintf("coef > %0.1f detected", coef_max))
      }
message("Cox divergent coefficients detection is finished")
      
# detection of (quasi-)complete separation
concordance_val <- tryCatch({
  suppressWarnings(summary(results_train$model)$concordance[1])
  }, error = function(e) NA)

message(sprintf("Detection of (quasi-)complete separation is finished ，concordance_val = %.4f", if (is.na(concordance_val)) NA else concordance_val))
      
if (!is.na(concordance_val) && concordance_val >= min_concord) {
        message(sprintf("Concordance >= %f detected ", min_concord))
      }

# val_result <- calculate_time_auc_cindex("Cox", fitted_model = results_train$model, df_test)
# val_result_df <- data.frame(
#   method = "Cox",
#   feature = "Clinical-Genom",
#   iAUC = val_result$iAUC,
#   c_index = val_result$c_index
# )
# kable_table_func(val_result_df, "cox_validation_results")

# Clinical-only COX
predictors_cli <- predictors[1:4]
results_train_cli <- fit_cox_model(predictors_cli, df)
cli_cox_result <- results_train_cli$coef_table
kable_table_func(cli_cox_result, "Clinical Cox Model Analysis Summary")
cli_cox_ph_test <- as.data.frame(results_train_cli$ph_test$table)
kable_table_func(cli_cox_ph_test, "Clinical Cox Model Proportional Hazards Test")
#val_result_cli <- calculate_time_auc_cindex("Cox", fitted_model = results_train_cli$model, df_test)

```

As seen in the above tables the Cox model shows that Grade2 is significant: HR ≈ 2.39 (p = 0.034) vs. the reference; grade3 is not significant.Several genes show strong associations (e.g., ge_202859_x_at HR ≈ 4.02; ge_220106_at HR ≈ 3.42; ge_217263_x_at protective, HR ≈ 0.34).Age, ER, tumor size are not statistically significant.Overall fit tests (LR/Wald/Score) are highly significant. The PH assumption is violated globally (p ≈ 1e−12) and for multiple covariates (e.g., age, grade, size, several genes). On the validation data set, the Time-weighted average AUC (iAUC) is `r val_result$iAUC` and the Concordance Index (C-index) is val_result\$c_index。However, the above results come from a single random train–test split; a comprehensive assessment requires cross-validation.

After identifying significant genes, we will fit two models—a Cox proportional hazards model and a Random Survival Forest—with training/validation conducted under bootstrap resampling.

## Random Survival Forest model

Random Survival Forest (RSF) is an extension of the Random Forest method to the survival analysis domain.It consists of many survival trees, built using bootstrap samples and random feature selection at each split. Splitting rule: Instead of maximizing information gain (e.g., Gini index or entropy in classification), survival trees use the log-rank test to find the feature and split point that best separate survival outcomes. For standard random forests, the leaf output is class label (classification) or numeric value (regression). In RSF, each terminal node (leaf) stores an estimated cumulative hazard function (CHF) (H(t)). For a new sample, each tree returns the CHF of the leaf the sample falls into, and the ensemble prediction is the average CHF across trees $H_{\text{avg}}(t)$ and convert to a survival function:\
$$S(t) = \exp(-H_{\text{avg}}(t))$$.

RSF outputs a survival curve, which gives the estimated probability of surviving beyond any time point (t). Besides individual survival prediction, RSF also supports variable importance analysis. With the library(randomForestSRC)  `rfsrc(Surv(t_dmfs, e_dmfs) ~ ., data = train_data, ntree = ntree)`can be used to build the RSF model.

This analysis utilized a random survival forest comprising 1,000 trees, with model training conducted through k-fold cross-validation, followed by selection of the optimal model for testing evaluation.Variable importance was determined using the full dataset.

```{r}
# Fit Random Survival Forest model with the same data as the cox model above
#result_rsf_train <- rsf_kfold_cv_best(data = df, K = 5, ntree = 1000)
#saveRDS(result_rsf_train, "result_rsf_train2025_09_19.rds")

result_rsf_train <- readRDS("result_rsf_train2025_09_19.rds")
rsf_fit_best <- result_rsf_train$best_model
rsf_fit_full <- rfsrc(Surv(t_dmfs, e_dmfs) ~ ., data = df, ntree = 1000)
rsf_v_importance <- data.frame(Variable = names(result_rsf_train$importance), 
                              Importance = result_rsf_train$importance) %>% 
  arrange(desc(abs(Importance))) 

kable_table_func(rsf_v_importance, "Variable Importance from Random Survival Forest model")
```

## Model Discrimination Assessment Metrics

### C-index (Concordance Index)

In this study, the discriminatory power of the survival model was assessed using the average C-index (Concordance Index) and average iAUC (Integrated Area Under the Curve) obtained through repeated cross-validation. The C-index is calculated by randomly selecting a pair of individuals $(i, j)$. The pair is considered concordant if: - Individual $i$ has an actual event time shorter than individual $j$, and - The model assigns a higher risk score (or shorter predicted survival time) to individual $i$. The C-index represents the proportion of concordant pairs among all comparable pairs:

$$
C = \frac{\text{Number of concordant pairs}}{\text{Number of comparable pairs}}
$$

Interpretation of Values: - $C = 0.5$ → Prediction is no better than random guessing - $C = 1.0$ → Perfect discriminatory ability - $0.7 \leq C < 0.8$ → Acceptable discrimination - $0.8 \leq C < 0.9$ → Excellent discrimination - $C \geq 0.9$ → Outstanding discrimination

### iAUC (Integrated Area Under the Curve)

The iAUC is the integrated form of time-dependent AUC. It measures a model's discriminatory ability at different time points and provides an overall summary score through integration (weighted averaging) over time. The calculation proceeds in two main steps:

i.  Compute time-dependent AUC ($t$): At a specific time point $t$, define two groups:

    -   The event group: individuals who experienced the event by time $t$.
    -   The non-event/survival group: individuals who remain event-free beyond time $t$ (including those censored after $t$).

    Calculate the standard AUC (Area Under the ROC Curve) for classifying these two groups at time $t$ based on the model's predicted risk scores. Since the risk scores are fixed, at each time point $t$ we simply reclassify the samples into the event group and the survival group. We then compare the risk scores to see whether the model correctly ranks those who experienced the event before $t$ ahead of those who survived beyond $t$.

ii. Integrate over all observed time points: The overall iAUC is obtained by integrating the time-dependent AUC($t$) function over the entire follow-up period $[0, \tau]$:

$$
\mathrm{iAUC} = \int_{0}^{\tau} \mathrm{AUC}(t)\, w(t)\, dt
$$ Where: - $\tau$ is the maximum observation time. - $w(t)$ is a weight function, often based on the distribution of event times or the size of the risk set at time $t$ to account for the varying precision of the estimate across time.

Interpretation of iAUC Values is same as the C-index.

Both the C-index and iAUC provide valuable insights into model performance: The C-index primarily assesses risk ordering but does not account for time-dependent dynamics in model performance which is addressed bythe iAUC. The combination of these metrics through repeated cross-validation ensures a robust assessment of the survival model's discriminatory power. The table below shows the C-index and iAUC of the COX and RSF models, which were built using both genetic and clinical data, along with the performance of models using clinical data alone, all evaluated on the test set.

```{r}

# evaluate the full Cox model 
result_valid <- calculate_time_auc_cindex(
        "Cox", 
        fitted_model = results_train$model, 
        df = clinical_cleaned_risk_test
      )

result_valid_cox_mc <- calculate_time_auc_cindex(
        "Cox", 
        fitted_model = results_train_cli$model, 
        df = clinical_cleaned_risk_test
      )
      
# evaluate the Random Survival Forest model
result_rsf_valid <- calculate_time_auc_cindex(
  "RSF",
  fitted_model = rsf_fit_full, #rsf_fit_best
  df = clinical_cleaned_risk_test
  )

# model comparision
comparison_df <- data.frame(
  Model = c("COX Clinical-Genomic", "RSF Clinical-Genomic", "COX Clinical-only"),
  iAUC = c(result_valid$iAUC, result_rsf_valid$iAUC, result_valid_cox_mc$iAUC),
  C_Index = c(result_valid$c_index, result_rsf_valid$c_index, result_valid_cox_mc$c_index),
  Sum_iAUC_Cindex = c(result_valid$iAUC, result_rsf_valid$iAUC, result_valid_cox_mc$iAUC) + c(result_valid$c_index, result_rsf_valid$c_index, result_valid_cox_mc$c_index)
)

kable(comparison_df, caption = "Optimal Iteration Performance in 100 Random Train-Test Splits")

#saveRDS(results_train$model, "best_cox_mm_model_1470.rds")
#saveRDS(results_train_cli$model, "best_cox_mc_model_1515.rds")
#saveRDS(rsf_fit_full, "best_rsf_model_mm_1489.rds")
```

## Botstrap cross validation using three separate feature sets

To compare the performances of the Cox proportional hazards model and the Random Survival Forest model, each algorithm were trained and evaluated using three separate feature sets: an exclusive clinical variable set, an exclusive gene expression variable set, and a combined set integrating both clinical and genomic features. Through bootstrap cross validations with replacement, we calculate the everage of C_index and iAUC.

```{r}
geo_accession_7390 <- clinical_cleaned_7390$geo_accession
expr_mat_7390 <- expr[, geo_accession_7390, drop = FALSE]

# res7390_3_model4 <- run_bootstrap_validation_3_model(expr_mat = expr_mat_7390,
#                                                      clinical_df = clinical_cleaned_7390,
#                                                      B = 100, 
#                                                      seed = 92000000, 
#                                                      min_epv = 5, 
#                                                      coef_max = 10, 
#                                                      min_concord = 0.97)

# Create a comparison dataframe
res7390_3_model4 <- readRDS("res7390_3_model4.rds")

model_comparison_mean <- data.frame(
  Model_Type = c("Clinical-only", "Genomic-only", "Clinical-Genomic Integrated"),
  Mean_Cox_iAUC = c(
    res7390_3_model4$summary_results_mc$mean_cox_iAUC,
    res7390_3_model4$summary_results_mg$mean_cox_iAUC,
    res7390_3_model4$summary_results_mm$mean_cox_iAUC
  ),
  Sd_Cox_iAUC = c(
    sd(res7390_3_model4$summary_results_mc$cox_iAUC),
    sd(res7390_3_model4$summary_results_mg$cox_iAUC),
    sd(res7390_3_model4$summary_results_mm$cox_iAUC)
  ),
  Mean_Cox_Cindex = c(
    res7390_3_model4$summary_results_mc$mean_cox_cidx,
    res7390_3_model4$summary_results_mg$mean_cox_cidx,
    res7390_3_model4$summary_results_mm$mean_cox_cidx
  ),
  Sd_Cox_Cindex = c(
    sd(res7390_3_model4$summary_results_mc$cox_cidx),
    sd(res7390_3_model4$summary_results_mg$cox_cidx),
    sd(res7390_3_model4$summary_results_mm$cox_cidx)
  ),
  Mean_RSF_iAUC = c(
    res7390_3_model4$summary_results_mc$mean_rsf_iAUC,
    res7390_3_model4$summary_results_mg$mean_rsf_iAUC,
    res7390_3_model4$summary_results_mm$mean_rsf_iAUC
  ),
  Sd_RSF_iAUC = c(
    sd(res7390_3_model4$summary_results_mc$rsf_iAUC),
    sd(res7390_3_model4$summary_results_mg$rsf_iAUC),
    sd(res7390_3_model4$summary_results_mm$rsf_iAUC)
  ),
  Mean_RSF_Cindex = c(
    res7390_3_model4$summary_results_mc$mean_rsf_cidx,
    res7390_3_model4$summary_results_mg$mean_rsf_cidx,
    res7390_3_model4$summary_results_mm$mean_rsf_cidx
  ),
  Sd_RSF_Cindex = c(
    sd(res7390_3_model4$summary_results_mc$rsf_cidx),
    sd(res7390_3_model4$summary_results_mg$rsf_cidx),
    sd(res7390_3_model4$summary_results_mm$rsf_cidx)
  )
)

# model_comparison_mean <- data.frame(
#   Model_Type = c("Clinical-only", "Genomic-only", "Clinical-Genomic Integrated"),
#   Mean_Cox_iAUC = c(
#     res7390_3_model4$summary_results_mc$mean_cox_iAUC,
#     res7390_3_model4$summary_results_mg$mean_cox_iAUC,
#     res7390_3_model4$summary_results_mm$mean_cox_iAUC
#   ),
#   Mean_Cox_Cindex = c(
#     res7390_3_model4$summary_results_mc$mean_cox_cidx,
#     res7390_3_model4$summary_results_mg$mean_cox_cidx,
#     res7390_3_model4$summary_results_mm$mean_cox_cidx
#   ),
#   Mean_RSF_iAUC = c(
#     res7390_3_model4$summary_results_mc$mean_rsf_iAUC,
#     res7390_3_model4$summary_results_mg$mean_rsf_iAUC,
#     res7390_3_model4$summary_results_mm$mean_rsf_iAUC
#   ),
#   Mean_RSF_Cindex = c(
#     res7390_3_model4$summary_results_mc$mean_rsf_cidx,
#     res7390_3_model4$summary_results_mg$mean_rsf_cidx,
#     res7390_3_model4$summary_results_mm$mean_rsf_cidx
#   )
# )

model_comparison_mean$Cox_Sum <- model_comparison_mean$Mean_Cox_iAUC + model_comparison_mean$Mean_Cox_Cindex
model_comparison_mean$RSF_Sum <- model_comparison_mean$Mean_RSF_iAUC + model_comparison_mean$Mean_RSF_Cindex

# Print the comparison table
#kable(model_comparison_mean, caption = "Table 14: Model Performance Comparision (Mean) from 100 Bootstrap Iterations") 

model_summary_df <- res7390_3_model4$summary_df %>%
  rename(Model_Type = model)
model_summary_df$Model_Type <- c(
  "Clinical-only",
  "Genomic-only", 
  "Clinical-Genomic Integrated"
)
#kable(model_summary_df, caption = "Table 15: Model Best Performance from 100 Bootstrap Iterations") 

model_summary_df <- model_summary_df[match(model_comparison_mean$Model_Type, 
                                           model_summary_df$Model_Type), ] 

merged_df_3_m <- model_comparison_mean %>%
  left_join(model_summary_df, by = "Model_Type") 
p_global <- results_train_cli$ph_test$table["GLOBAL", "p"]
merged_df_3_m$Global_PH_Test_p <-c(p_global, NA, NA)

cn <- colnames(merged_df_3_m)
cn_new <- gsub(" ", "_", tools::toTitleCase(gsub("_", " ", cn)))
colnames(merged_df_3_m) <- cn_new

show_col <- c("Model_Type",
              "Mean_Cox_iAUC",
              "Sd_Cox_iAUC",
              "Mean_Cox_Cindex",
              "Sd_Cox_Cindex",
              "Mean_RSF_iAUC",
              "Sd_RSF_iAUC",
              "Mean_RSF_Cindex",
              "Sd_RSF_Cindex"
              #"Cox_Sum",
              #"RSF_Sum"
              #"Rsf_Predictors"
              #"Best_Performance",
              #"Best_Iteration"
              #"Best_Seed"
              #"Best_Method",
              #"Global_PH_Test_p"
              )

merged_df_3_m2 <- merged_df_3_m %>% select(show_col)
kable(merged_df_3_m2, digit = 2,  caption = "3 Model Performance Comparision from 100 Bootstrap Iterations") 
```

Across 100 bootstrap resamples, the Clinical-Genomic Integrated Cox model yielded a mean iAUC `r merged_df_3_m2$Mean_Cox_iAUC[3]` and C-index `r merged_df_3_m2$Mean_Cox_Cindex[3]` with standard deviations around 0.09 indicating no performance advantage over a random guess. Similarly, the other models exhibited comparable performance. Below, we proceed with dimensionality reduction of the gene expression data followed by modeling using the Cox Proportional Hazards (Cox PH) approach.

##PCA Cox: Cox model built on Principal Component Analysis (PCA) results

PCA is an unsupervised, linear method that rotates the data to new, orthogonal axes (principal components) that capture the largest possible variance first. Each sample gets scores on these components; each gene has loadings that show how strongly it contributes. PCA is typically computed by an eigen-decomposition of the covariance (or correlation) matrix, or via singular value decomposition (SVD).

PCA reduces dimensionality, mitigates multicollinearity, and can denoise high-dimensional data. Keeping only the top𝑘components often preserves most of the signal while discarding noise, which can improve model stability and computation. In gene expression studies, PCA can compress \~20,000 genes into a few dozen components that retain most variation, yielding more stable estimates and faster computation in downstream analyses.

In this study, we reduced the dimensionality of gene-expression data via PCA fitted on the training set and used the resulting principal component (PC) scores as predictors in Cox proportional hazards (PH) models. The validation set was projected using the training-set centering, scaling, and rotation (loadings). Model performance was summarized over 100 bootstrap resamples. When selecting the number of PCs, we enforced an events-per-variable (EPV) rule of 3. The goal is to ensure that the retained principal components explain at least 50% of the variance in the original data.

```{r}

expr_df <- as.data.frame(t(expr))
#combine gene with clinical data
clinical_t_e <- clinical_cleaned_7390 %>% select(geo_accession, t_dmfs, e_dmfs)
merged_df <- expr_df %>%
  rownames_to_column("geo_accession") %>%
  left_join(clinical_t_e, 
            by = "geo_accession") %>%
  column_to_rownames("geo_accession")

n_runs = 100
# pca_cox_result <- bootstrap_train_cox_with_pca(data = merged_df,
#                                            clinical_df = clinical_cleaned_7390,
#                                            n_runs = n_runs,
#                                            seed = 12395 )

#EPV=3 model = TRUE in trainning
pca_cox_result <- readRDS("PCA_COX_result_model-inclusive_9_18_2.rds") 
pca_avg_results <- pca_cox_result$avg_results
pca_best_results_df <- pca_cox_result$best_results_df

kable_table_func(pca_avg_results, "Cox model built on PCA CV result (n_runs = 100)")
```

Exploratory analyses suggested between-hospital heterogeneity in survival (p = 0.10). Motivated by this, we also fit a Cox mixed-effects model (coxme) with a hospital-level random effect, but it did not improve validation performance. The mean C-index performance fell below random chance levels. This sub-random performance is presumably attributable to severe sample insufficiency after incorporating random effects into the model.

```{r}
Mean_Cox_iAUC   <- pca_avg_results[1, 2]   # C-index
Mean_Cox_Cindex <- pca_avg_results[3, 2]   # iAUC
best_perf       <- pca_best_results_df[5, 2] # Total Score
best_iter       <- pca_best_results_df[1, 2] # Best Run ID
best_seed       <- pca_best_results_df[2, 2] # Best Seed
p_global_pca    <- pca_best_results_df[8, 2]

new_row_pca <- data.frame(
  Model_Type        = "PCA-Genomic-only",
  Mean_Cox_iAUC     = Mean_Cox_iAUC,
  Mean_Cox_Cindex   = Mean_Cox_Cindex,
  Mean_RSF_iAUC     = NA,
  Mean_RSF_Cindex   = NA,
  Cox_Sum           = Mean_Cox_iAUC + Mean_Cox_Cindex,
  RSF_Sum           = NA,
  Rsf_Predictors    = NA,
  Best_Performance  = best_perf,
  Best_Iteration    = best_iter,
  Best_Seed         = best_seed,
  Best_Method       = "COX",
  Global_PH_Test_p  = p_global_pca
)

merged_df_4_m <- rbind(merged_df_3_m, new_row_pca)
merged_df_4_m2 <- merged_df_4_m %>% select(show_col)
kable(merged_df_4_m2, digit = 2,  caption = "4 Model Performance Comparision") 

```

It can be observed that after 100 iterations of bootstrapped cross-validation, the Cox model built on PCA principal components achieved mean C-index and iAUC values of `r merged_df_4_m2$Mean_Cox_Cindex[4]` and `r merged_df_4_m2$Mean_Cox_iAUC[4]`, respectively. Again this indicates suboptimal predictive power but they are slightly better than other models.

## VAE Cox: Cox model built on VAE latent representation

A Variational Autoencoder (VAE) is a type of deep generative model that learns to represent high-dimensional data, such as gene expression profiles, in a lower-dimensional latent space.In this project this method was also explored, with the resulting latent variables incorporated into Cox regression modeling.

VAE consists of two parts: an encoder, which compresses the input data into a set of latent variables, and a decoder, which reconstructs the data from these variables. Unlike a standard autoencoder, a VAE assumes a probabilistic distribution for the latent variables (typically Gaussian), allowing it to generate new data samples by sampling from this distribution. This property makes VAEs useful not only for dimensionality reduction but also for capturing complex patterns in the data and creating smooth latent representations that can be used in downstream analyses, such as survival modeling.

```{r}
# prepare data
library(torch)
source("qmd_functions_vae.R")
train_expr_df <- t(train_expr_filtered)
test_expr_df <- t(test_expr_scaled)
x_data <- torch_tensor(train_expr_df, dtype = torch_float()) 
x_val_data <- torch_tensor(test_expr_df, dtype = torch_float())
latent_dims = c(4,6,10, 11, 12, 13, 14, 15, 16, 20, 30)

# find the best latent dim
# best_2 <- find_best_latent_dim(
#   latent_dims = latent_dims,
#   x_data = x_data,
#   x_val_data = x_val_data,
#   hidden_dim = 256
# )
# results <- best_2$results
# best_latent_dim <- best_2$best_latent_dim
# best_overall_loss <- best_2$best_overall_loss
# saveRDS(best_2, "best_latent_dim30.rds")

best2 <- readRDS("best_latent_dim30.rds")
results <- best2$results
best_latent_dim <- best2$best_latent_dim
best_overall_loss <- best2$best_overall_loss

# Plot
if (length(results) > 0) {
  latent_dims <- sapply(results, function(x) x$latent_dim)
  val_losses <- sapply(results, function(x) x$best_val_loss)
  
  plot(latent_dims, val_losses, type = "b", pch = 19, col = "blue",
       xlab = "Latent Dimension", ylab = "Best Validation Loss",
       main = "VAE Performance vs Latent Dimension")
  points(best_latent_dim, best_overall_loss, pch = 19, col = "red", cex = 1.5)
  legend("topright", legend = "Best", pch = 19, col = "red")
}


best_latent_dim2 <- min(max_vars_allowed,  best_latent_dim)


```

In this project, VAE models were first trained on the training dataset across multiple latent dimensions `r latent_dims`, and the reconstruction loss + Kullback–Leibler (KL) divergence was evaluated on an independent test set. The latent dimension `r best_latent_dim2` yielding the lowest loss was then selected as the basis for subsequent Cox regression modeling.

In the next step a VAE model was first constructed using the training data, from which the latent representations of training data were extracted. A Cox model based solely on these latent variables was then built. The validation data were subsequently transformed into latent representations with the VAE model and used to evaluate the Cox model.Fifty iterations of cross-validation with replacement were conducted.

It is worth noting that, to reduce computational load in the VAE pipeline, we preselected genes prior to training by (i) applying median absolute deviation (MAD) filtering to remove low-variability genes, (ii) conducting univariate Cox screening, and (iii) pruning highly correlated genes. We then standardized the training set (with the same centering/scaling applied to validation data). These preprocessing steps were not applied in the PCA-based workflow.

```{r}
# res <- run_vae_cox_iterations(
#   expr          = expr,                   
#   clinical_df   = clinical_cleaned_7390,  
#   n_iterations  = 50,
#   seed          = 2025091520,
#   latent_dim    = 11,
#   hidden_dim    = 128,
#   epochs        = 800,
#   patience      = 50,
#   lr            = 0.001,
#   mad_quantile  = 0.25,
#   uni_cox_p     = 0.01,
#   corr_cutoff   = 0.90,
#   save_dir      = "best_models_dim11"           # NULL not save
# )

#saveRDS(res, "run_vae_cox_iterations_seed_2025091520.rds")
res = readRDS("run_vae_cox_iterations_seed_2025091520.rds")
best_cox_model_vae <- readRDS("G:/clinical_data_science/final_project_github/BreastCancerRecurrencePrediction/best_models_dim11/best_cox_model.rds")
ph_test <- cox.zph(best_cox_model_vae)
global_p_vae <- ph_test$table["GLOBAL","p"]

vae_best_iteration    <- res$best_iteration
vae_best_score        <- res$best_score
vae_iteration_results <- res$iteration_results
 
# # summary
iter_df     <- do.call(rbind, lapply(vae_iteration_results, as.data.frame))
mean_c_index <- mean(iter_df$c_index, na.rm = TRUE)
mean_iAUC    <- mean(iter_df$iAUC,   na.rm = TRUE)
sd_c_index   <- sd(iter_df$c_index,  na.rm = TRUE)
sd_iAUC      <- sd(iter_df$iAUC,     na.rm = TRUE)

vae_avg_metrics <- list(mean_c_index = mean_c_index,
                    sd_c_index   = sd_c_index,
                    mean_iAUC    = mean_iAUC,
                    sd_iAUC      = sd_iAUC,
                    n_iterations = nrow(iter_df))

new_row_vae <- data.frame(
  Model_Type        = "VAE-Genomic-only",
  Mean_Cox_iAUC     = mean_iAUC,
  Mean_Cox_Cindex   = mean_c_index,
  Mean_RSF_iAUC     = NA,
  Mean_RSF_Cindex   = NA,
  Cox_Sum           = mean_iAUC + mean_c_index,
  RSF_Sum           = NA,
  Rsf_Predictors    = NA,
  Best_Performance  = vae_best_score,
  Best_Iteration    = vae_best_iteration,
  Best_Seed         = 2025091520 + vae_best_iteration,
  Best_Method       = "COX",
  Global_PH_Test_p  = global_p_vae
)

merged_df_5_m <- rbind(merged_df_4_m, new_row_vae)

kable(merged_df_5_m, caption = "Table 16: Model Performance Comparision") 

```

Based on the test results, none of the models demonstrate satisfactory discriminative performance on the test set. The average scores are around 0.5, only slightly better than random classification. One reason for this is the small sample size, as well as random factors such as hospital-related effects. In addition, the algorithms and overall approach still require further improvement.

Among all models, the Cox model constructed after applying PCA for dimensionality reduction on the gene expression data performed better on average than the others. In its best iteration, the combined iAUC + C-index reached 1.60, but the proportional hazards (PH) test was not passed. Nevertheless, since the project requires selecting one model as the predictive model for future use, we have chosen this model as the predictive model.

## Model application and Prediction

In order to perform personalized survival prediction for new patients using a pre-trained PCA-COX model, the genomic data is needed to be standardized using the same center and scale parameters from the original training set to maintain consistency. Then the standardized data is transformed into principal component space using the rotation matrix from the trained PCA model, retaining only the most significant components based on the original variance threshold.

The transformed data creates the input for COX regression prediction. The pre-trained COX model then generates three key predictions: a linear risk score indicating relative risk, a hazard ratio representing the multiplicative increase in risk compared to the baseline, and most importantly, a personalized survival probability estimate specifically at the patient's own observed follow-up time.

```{r}
set.seed(330)
new_pat_id <- sample(1: nrow(merged_df), 1) # 140
new_patient_data <- merged_df[new_pat_id,]
pca_cox_prediction <- predict_pca_cox(new_pat = new_patient_data, pca_cox_result)
kable(pca_cox_prediction, caption = "PCA Cox prediction VS real survive time") 
```

## Discussion

We evaluated multiple Cox proportional hazards (PH) models and random survival forests (RSF) on clinical and gene-expression data, and also built Cox models on PCA and VAE–derived representations. All models performed suboptimally. Likely contributors include: (1) limited sample size and few events—low events-per-variable (EPV) leading to unstable/overfit estimates; (2) violations of the PH assumption; (3) the $p \gg n$ regime and multicollinearity in gene data, which harm stability and generalization; (4) suboptimal hyperparameters (e.g., for VAE and RSF), suggesting a grid search over latent_dim, hidden_dim, ntree, etc.; and (5) weak signal—clinical/genomic predictors may have limited prognostic value, motivating alternative biological stratifications or data types.

For follow-up, one could fit RSF on PCA scores and on VAE latent variables. The relatively better performance of PCA may reflect preprocessing choices—for example, the absence of univariate Cox/LASSO pre-screening and/or omission of scaling before PCA in some configurations. These steps warrant careful reconsideration to avoid degrading signal representation or injecting unnecessary noise.

## Conclusion

Genomic data are increasingly important for cancer prognosis. Using gene-expression and clinical data from the GEO database (GSE7390), we conducted a survival analysis of breast cancer and compared Cox proportional hazards (PH) models with Random Survival Forests (RSF) across different feature representations. Missing clinical values were handled via multiple imputation by chained equations (MICE). To address high dimensionality in the genomic data, we explored PCA and VAE for dimensionality reduction, L1-penalized Cox, median absolute deviation (MAD) filtering, and correlation pruning. All preprocessing and feature selection were fit on the training set and then applied to the validation set to prevent data leakage. Model performance was evaluated using the C-index and the integrated time-dependent AUC (iAUC). Training and testing employed repeated bootstrap resampling to ensure robust, less-biased estimates. The model ultimately chosen for prospective use was a Cox PH model built on PCA-derived principal components. Although validation performance did not reach a clearly discriminative level, the study provides a useful baseline and guidance for future refinement.
