---
title: "Comparative Modeling Approaches for Breast Cancer Prognosis --- From Gene Selection with LASSO to PCA and VAE-Based Dimensionality Reduction"
author: "Baodong Zhang"
editor: visual
format:
  pdf:
    documentclass: article
    fontsize: 6pt
echo: true
warning: false
message: false
fig-width: 6
fig-height: 4
fig-align: center
geometry: "top=1in, bottom=1in, left=1in, right=1in"
header-includes:
  - \usepackage[table,xcdraw]{xcolor}
  
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,       
  eval = TRUE,       
  warning = FALSE,   
  message = FALSE,   
  error = FALSE,     
  results = 'markup' 
  )
library(dplyr)
library(survival)
library(survminer)
library(glmnet)
library(tibble)
library(kableExtra)
library(mice)
library(randomForestSRC)
library(riskRegression) # iAUC cindex
library(caret)
library(pec) # iAUC cindex
source("qmd_functions.R")
source("Project_functions.R")
options(scipen = 999)
```

## Summarize

This project explored multiple approaches for predicting distant recurrence-free survival in breast cancer using genomic data. Starting with over 20,000 genes, univariate Cox regression identified 800+ candidates, which were further reduced to 40 genes using LASSO-Cox. The study compared model performance across (i) clinical variables alone, (ii) clinical variables plus 40 genes, and (iii) 40 genes alone, using both Cox and random survival forest models. Dimensionality reduction methods were also tested, including PCA and a variational autoencoder (VAE) applied to the 800 genes, with latent variables subsequently modeled via Cox regression. Extensive bootstrap cross-validation was employed to evaluate models by mean iAUC and c-index across iterations. Data preprocessing included MICE imputation for missing clinical values, z-score normalization of gene data, and strict separation of training and validation sets to prevent data leakage. Through 100 bootstrap cross validations with replacement, it was found that the COX model with pure clinic variables achieved the optimal value of 0.61 in terms of average iAUC. In 100 iterations of the COX model with pure clinic variables, the model with the best performance of iAUC+C-index was selected as the final model for future prediction.

## Introduction

In the study of breast cancer, survival analysis has always been a meaningful topic. Identifying the key factors that affect patients' survival time and quality of life is a prerequisite for developing effective treatment methods. This project attempts to use the Cox Lasso model to screen gene expression that affects survival time, and establish a model together with clinical variables to predict the future survival time of patients.

## Data Source

The data for this project were sourced from the internationally renowned NCBI GEO Database (Gene Expression Omnibus) under the accession number GSE7390, which includes a total of 198 breast cancer patients. All patients were lymph node-negative.

Clinical variables include age, tumor grade, estrogen receptor status (ER status),and tumor size.

The gene expression data were obtained using microarray technology and comprise 22,283 variables. The target variables include: Recurrence status: A binary variable indicating whether the patient experienced disease recurrence (yes/no). Time to recurrence: The duration from treatment to the occurrence of disease recurrence, measured in days. For patients who did not experience recurrence, the last known recurrence-free time was recorded (i.e., right-censored data).

```{r}

expr <- readRDS("expr_7390.RDS")
clinical <- readRDS("clinical_7390.rds")
#subset_tibble <- as_tibble(expr[1:5, 1:5], rownames = "Probe_ID")

# process clinical data
clinical$time <- as.numeric(clinical$`t.dmfs:ch1`)
clinical$status <- as.numeric(clinical$`e.dmfs:ch1`)

clinical_selected <- clinical[, c(
  "geo_accession",
  "t.dmfs:ch1",
  "e.dmfs:ch1",
  "age:ch1",
  "er:ch1",
  "grade:ch1",
  "size:ch1",
  "node:ch1",
  "hospital:ch1", "samplename:ch1", "filename:ch1"
)]

original_names <- colnames(clinical_selected)
clean_names <- gsub(":ch1", "", original_names)
clean_names <- gsub("\\.", "_", clean_names)
colnames(clinical_selected) <- clean_names

clinical_cleaned <- as.data.frame(lapply(clinical_selected, function(x) {
  x_num <- suppressWarnings(as.numeric(as.character(x)))
  if (all(is.na(x_num))) x else x_num
}))

# convert to vectors
# for (i in 4:ncol(clinical_cleaned)) {
#   if (length(unique(clinical_cleaned[, i])) <= 4) {
#     clinical_cleaned[, i] <- factor(clinical_cleaned[, i])
#   }
# }

clinical_cleaned_7390 <- clinical_cleaned %>%
  select(geo_accession, t_dmfs, e_dmfs, age, er, grade, size, hospital)


kable_table_func(subset_tibble, "Table 1: Subset of Gene Expression Matrix (Top 5 Probes and Samples)")
kable_table_func(head(clinical_cleaned_7390, 5), "Table 2: Subset of Clinical Data (Top 5 Probes and Samples)")

```

## Data Preprocessing and Exploratory Data Analysis (EDA)

Clean clinical data: handle missing values and outliers

```{r}

numeric_vars <- names(clinical_cleaned_7390)[sapply(clinical_cleaned_7390, is.numeric)]
generate_box_plots(clinical_cleaned_7390, numeric_vars)

```

The box plot of clinical data shows no outliers.

```{r}

# Detect and define factor variables appropriately:
for (i in 4:7) {
 if (length(unique(clinical_cleaned_7390[, i])) <= 4) {
   clinical_cleaned_7390[, i] <- factor(clinical_cleaned_7390[, i])
 }
}

na_info <- colSums(is.na(clinical_cleaned_7390))
na_info <- na_info[na_info > 0]

# 以数据框形式显示
na_df <- data.frame(
  Column = names(na_info),
  NA_Count = as.numeric(na_info),
  row.names = NULL
)
na_df$Class <- "factors level > 2"
na_df$Mice_Method <- "polyreg"
kable_table_func(na_df, "Table 3: NA Audit Report")

```

Handling missing values in clinical data.

The clinical dataset exhibited missingness exclusively in the grade column, with two incomplete observations. As grade represents an ordinal factor variable with three levels, multiple imputation was performed using the proportional odds logistic regression ("polr") method within the MICE framework. The final imputed values were derived through majority voting across the multiply imputed datasets.

To prevent data leakage, the MICE algorithm was trained solely on the training subset. The validation set was subsequently imputed using the parameters estimated from the training data, thereby preserving the validation set's integrity for unbiased model performance assessment.

```{r}

clinical_cleaned_7390_imputed <- impute_missing_value(clinical_cleaned_7390, selected_col= c("age", "er", "grade", "size"), missed_col = "grade")
hospital_counts <- table(clinical_cleaned_7390_imputed$hospital)

```

```{r}

fit <- survfit(Surv(t_dmfs, e_dmfs) ~ hospital, data = clinical_cleaned_7390_imputed)
ggsurvplot(fit, data = clinical_cleaned_7390_imputed, pval = TRUE) 


# library(coxme)
# coxph_model <- coxme(Surv(time = t_dmfs, event = e_dmfs) ~ age + er + grade + size + (1 | hospital), data = clinical_cleaned_7390)
# summary(coxph_model)

coxph_pur <- coxph(Surv(time = t_dmfs, event = e_dmfs) ~ age + er + grade + size , data = clinical_cleaned_7390_imputed)
summary(coxph_pur)


```

There are `r length(unique(clinical_cleaned_7390$hospital))` hospitals in the clinical dataset. The Kaplan-Meier survival curves generated by ggsurvplot demonstrate partially overlapping patterns across these hospitals, and the log-rank test yields a p-value of 0.1. While this suggests some degree of heterogeneity in survival outcomes among different hospital sites, the p-value \> 0.05 indicates that the observed differences are not statistically significant at the conventional alpha level of 0.05. Given the limited sample size and the lack of statistical significance, the random effects associated with hospital variation will not be considered as a primary focus in subsequent analyses.

```{r}

max(expr)
min(expr)
row_means <- rowMeans(expr, na.rm = TRUE)
row_sds <- apply(expr, 1, sd, na.rm = TRUE)

summary(row_means)
summary(row_sds)

```

Handling gene data。

```{r}
set.seed(92000028)
expr_mat <- expr
clinical_df <- clinical_cleaned_7390
data_prepare_result <- bootstrap_sampl_split_miss_vale_imputation(expr_mat, clinical_df)
train_expr <- data_prepare_result$train_expr
test_expr <- data_prepare_result$test_expr
train_clinical <- data_prepare_result$train_clinical
test_clinical <- data_prepare_result$test_clinical
events_train <- data_prepare_result$events_train

# quality check
test_clinical_unimputed = clinical_df %>% filter( geo_accession %in% test_clinical$geo_accession)
setdiff(test_clinical_unimputed, test_clinical)
setdiff(test_clinical, test_clinical_unimputed)
# MAD filter
quantile_cutoff = 0.25
filtered_data <- filter_genes_by_mad(train_expr, test_expr, quantile_cutoff)
train_expr2 <- filtered_data$train_expr
test_expr2 <- filtered_data$test_expr

```

The value of gene expression ranges from `r min(expr)` to `r max(expr)` and has multiple decimal places. This range is typical data converted by Log2.

Through bootstrap sampling with replacement, train-test splitting was performed alongside multiple imputation of clinical data. For genomic data, Median Absolute Deviation (MAD) filtering was applied to remove genes with minimal expression variation in the training set. Genes demonstrating the highest `r 1-quantile_cutoff` of expression variability are retained for subsequent analysis. Gene quantity was reduced from `r nrow(expr_mat)` to `r nrow(train_expr2)` after preprocessing. The validation set retained the identical gene subset to ensure analytical consistency and prevent data leakage.

```{r}
kable_table_func(train_clinical[1:5,], "Table 4: Training set samples for clinical data")
kable_table_func(train_expr2[1:5, 1:5], "Table 5: Training set samples for gene expression data")
```

Performinf univariate cox regression to select significant genes:

The Cox proportional hazards model is a semi-parametric regression model used in survival analysis to examine the relationship between survival time and one or more predictor variables. The model specification is:

$$
h(t|X) = h_0(t) \times \exp(\beta_1X_1 + \beta_2X_2 + \cdots + \beta_pX_p)
$$

where:

-   $h(t|X)$ is the hazard function at time $t$ given covariates $X$
-   $h_0(t)$ is the baseline hazard function\
-   $\exp(\beta_iX_i)$ represents the hazard ratio for each covariate

```{r}
    sig_gene_df <- batch_univariate_cox_regression(train_expr2, train_clinical, p_value = 0.05)
    message(sprintf("Univariate cox regression finished, significant genes = %d", if (is.null(sig_gene_df)) 0 else nrow(sig_gene_df)))

```

Univariate Cox proportional hazards regression was performed individually for each gene, with significantly associated genes (Wald test p-value \< 0.05) subsequently identified and compiled into a results table.

```{r}

sig_gene_df <- sig_gene_df %>%
  select(gene, everything()) %>%
  remove_rownames()

kable_table_func(head(sig_gene_df, 5), "Table 6: Univariate_Cox_Significant_Genes_Table (top 5)")
```

Conducted LASSO Cox proportional hazards regression for further refinement of gene selection. Lasso (Least Absolute Shrinkage and Selection Operator) Cox regression is a regularized version of the Cox proportional hazards model that incorporates L1 penalty for simultaneous variable selection and coefficient estimation in survival analysis.

### Mathematical Formulation

The objective function combines partial likelihood with L1 penalty:

$$
\min_{\beta} \left[ -\frac{1}{n} \ell(\beta) + \lambda \sum_{j=1}^{p} |\beta_j| \right]
$$

**where:**

-   `$\ell(\beta)$` = partial likelihood of the Cox model
-   `$\lambda$` = regularization parameter controlling shrinkage
-   `$\sum_{j=1}^{p} |\beta_j|$` = L1 penalty term

The Lasso Cox survival modeling is to evaluate the prognostic impact of different genes/clinical variables and the relative strength of their effects. It is essential to compare effect sizes (hazard ratios) on a common scale. Therefore, Z-score normalization was applied individually to each gene and numerical clinical variable. Additional feature selection was performed using remove_high_corr_genes function with a correlation coefficient cutoff of 0.90 to address multicollinearity among predictors. To similarly prevent information leakage, the validation set was standardized utilizing the mean and standard deviation parameters calculated solely from the training dataset.

```{r}

# prepare data for Lasso Cox regression
significant_gene <- sig_gene_df$gene
train_expr_scaled <- standardize_with_train(gene_mat_train = train_expr2,
                                                gene_mat_valid = train_expr2,
                                                significant_gene = significant_gene)


train_expr_filtered <- remove_high_corr_genes(train_expr_scaled, cutoff = 0.90)
message("Correlation-based filtering completed: train_expr_filtered dimensions = ", 
        paste(dim(train_expr_filtered), collapse = " x "))

significant_gene2 <- rownames(train_expr_filtered)
message(sprintf("Successfully extracted filtered gene set, significant_gene2 contains %d features", 
                length(significant_gene2)))

# Normalize test set gene expression data using training set parameters
test_expr_scaled <- standardize_with_train(gene_mat_train = train_expr2,
                                           gene_mat_valid = test_expr2,
                                           significant_gene = significant_gene2)

kable_table_func((test_expr_scaled %>% 
   as.data.frame() %>% 
   rownames_to_column("gene"))[1:5, 1:5], "Table 8: Scaled/aligned test set samples for gene expression data")

# Normalized clinical data
columns_to_remove <- c("e_dmfs", "t_dmfs", "geo_accession", "hospital")
numeric_columns <- train_clinical %>%
      select(where(is.numeric)) %>%
      colnames
numeric_columns <- setdiff(numeric_columns, columns_to_remove)

train_clinical_scaled <- standardize_with_train_clinical(train_clinical,
                                                             train_clinical,
                                                             scale_cols = numeric_columns)
test_clinical_scaled <- standardize_with_train_clinical(train_clinical,
                                                            test_clinical,
                                                            scale_cols = numeric_columns)

```

The LASSO Cox regression procedure incorporated repeated cross-validation to enhance the robustness of gene selection across multiple iterations.

```{r}

gene_freq_df <- repeat_cv_lasso_cox(train_expr = train_expr_filtered,
                                        train_clinical,
                                        significant_gene_vec = significant_gene2,
                                        repeats = 5,
                                        nfolds = 10,
                                        alpha = 1)
message(sprintf("LASSO Cox finished"))
```

```{r}
min_epv <- 5 
min_freq <- 0.8
```

The resulting genes were prioritized based on selection frequency and mean coefficient magnitude, with a retention threshold set at ≥ `r min_freq` frequency.

Regarding sample size considerations, Peduzzi et al. (1995, J Clin Epidemiol) initially proposed a conservative requirement of 10 events per variable (EPV) for Cox regression. Vittinghoff & McCulloch (2007, Am J Epidemiol) later demonstrated that 5–9 EPV may be sufficient for many analytical scenarios. Consistent with the latter guidelines and considering the cohort size limitations, an EPV threshold of `r min_epv` was implemented for additional feature selection.

```{r}

gene_freq_df_best <- gene_freq_df %>%
      filter(freq >= min_freq)

max_vars_allowed <- floor(events_train / min_epv)

selected_gene_df <- gene_freq_df[1:min(nrow(gene_freq_df_best), max_vars_allowed), ] %>%
      mutate(coef = mean_coef)
#Cross_Validation_Consensus_Genes_Frequency_Threshold_80_EPV_5

kable_table_func(selected_gene_df, "Table 9: Multi_Iteration_LASSO_Genes_Selection_Frequency_0.8_EPV_5")


```

The risk score for patient will be calculated as:

The risk score is calculated as:

$$
\text{risk score} = \beta_1 x_1 + \beta_2 x_2 + \dots + \beta_i x_i + \dots + \beta_p x_p
$$ where: - $\beta_i$ represents the coefficient for each variable - $x_i$ represents the value of each predictor variable - $p$ is the total number of variables in the model

```{r}
result <- compute_risk_score_train_2(
      gene_mat_scaled  = train_expr_filtered,
      significant_vars_df = selected_gene_df,
      clinical_cleaned = train_clinical_scaled
      )

clinical_cleaned_risk_train <- result$merged_df
median_cut_train <- result$median_cut

clinical_cleaned_risk_test <- compute_risk_score_test_2(
  gene_mat_scaled = test_expr_scaled,
  significant_vars_df = selected_gene_df,
  clinical_cleaned = test_clinical_scaled,
  median_cut_train)
    

clinical_cleaned_risk_test$group_var <- as.factor(clinical_cleaned_risk_test[["risk_group"]])
  
surv_obj <- Surv(time = clinical_cleaned_risk_test$t_dmfs, 
                   event = clinical_cleaned_risk_test$e_dmfs)
  
  
# Fit survival model
fit <- survfit(surv_obj ~ risk_group, data = clinical_cleaned_risk_test)
  
# Create survival plot
ggsurvplot(
  fit,
  data = clinical_cleaned_risk_test,
  pval = TRUE,
  conf.int = TRUE,
  risk.table = TRUE,
  xlab = "Days",
  ylab = "Survival Probability",
  palette = "Set1"
  )
  
surv_diff <- survdiff(surv_obj ~ group_var, data = clinical_cleaned_risk_test)
p_value <- 1 - pchisq(surv_diff$chisq, length(surv_diff$n) - 1)
cat("Log-rank Kaplan-Meier test p-value:", p_value, "\n")

```

Following the identification of significant genes, subsequent modeling will employ two distinct algorithms: a Cox proportional hazards model and a Random Survival Forest model.

Cox proportional hazards model

During the fitting of a Cox model, for instance, if a specific gene expression is present in all patients who died within the first year but absent in all patients who survived beyond five years, the model will assign an extremely large coefficient to this variable, resulting in an exceptionally high concordance index. To account for the potential presence of (quasi-)complete separation and divergent coefficients, thresholds using `min_concord` and `coef_max` have been implemented . If the model's overall concordance exceeds `min_concord` (e.g., 0.97) **or** if any coefficient in the model is greater than `coef_max` (e.g., 10), the model will be rejected.

```{r}

columns_to_remove2 <- c("risk_group", "risk_score", "e_dmfs", "t_dmfs") #, 

character_columns <- clinical_cleaned_risk_train %>%
  select(where(is.character)) %>%
  colnames()

# columns_with_digits <- clinical_cleaned_risk_train %>%
#   select(matches("\\d")) %>%
#   names()
      
predictors <- colnames(clinical_cleaned_risk_train) %>%
  setdiff(c(columns_to_remove2, character_columns)) # , columns_with_digits

df <- clinical_cleaned_risk_train %>%
  select(c(predictors, "t_dmfs", "e_dmfs"))
# fit cox
results_train <- fit_cox_model(predictors, df)

full_cox_result <- results_train$coef_table

kable_table_func(full_cox_result, "Table 10: Full Cox Model Analysis Summary")

full_cox_ph_test <- results_train$ph_test

kable_table_func(full_cox_ph_test, "Table 11: Full Cox Model Proportional Hazards")

#print(full_cox_ph_test)

ph_plots <- ggcoxzph(results_train$ph_test)
print(ph_plots[[1]])
print(ph_plots[[2]])
print(ph_plots[[3]])
print(ph_plots[[4]])
print(ph_plots[[5]])

coef_max <- 10
min_concord <- 0.97


# detection of divergent coefficients in the Cox model
if (any(abs(coef(results_train$model)) > coef_max)) {
  message(sprintf("coef > %0.1f detected", coef_max))
      }
message("Cox divergent coefficients detection is finished")
      
# detection of (quasi-)complete separation
concordance_val <- tryCatch({
  suppressWarnings(summary(results_train$model)$concordance[1])
  }, error = function(e) NA)

message(sprintf("Detection of (quasi-)complete separation is finished ，concordance_val = %.4f", if (is.na(concordance_val)) NA else concordance_val))
      
if (!is.na(concordance_val) && concordance_val >= min_concord) {
        message(sprintf("Concordance >= %f detected ", min_concord))
      }

# Clinical-only COX
predictors_cli <- predictors[1:4]
results_train_cli <- fit_cox_model(predictors_cli, df)
```

Random Survival Forest model

Random Survival Forest (RSF) is an extension of the Random Forest method to the survival analysis domain.It consists of many survival trees, built using bootstrap samples and random feature selection at each split. Splitting rule: Instead of maximizing information gain (e.g., Gini index or entropy in classification), survival trees use the log-rank test to find the feature and split point that best separate survival outcomes. Leaf output: - Standard random forests → class label (classification) or numeric value (regression). RSF → cumulative hazard function (CHF), Ensemble prediction: - Each tree provides a CHF (H(t)). The forest averages them to obtain $H_{\text{avg}}(t)$ and convert to a survival function:\
$$S(t) = \exp(-H_{\text{avg}}(t))$$ RSF outputs a survival curve, which gives the estimated probability of surviving beyond any time point (t). Besides individual survival prediction, RSF also supports variable importance analysis. With the library(randomForestSRC), `rfsrc(Surv(t_dmfs, e_dmfs) ~ ., data = train_data, ntree = ntree)`can be used to build the RSF model.

The analysis utilized a random survival forest comprising 1,000 trees, with model training conducted through k-fold cross-validation, followed by selection of the optimal model for testing evaluation.Variable importance was determined using the full dataset.

```{r}
# Fit Random Survival Forest model
result_rsf_train <- rsf_kfold_cv_best(data = df, K = 5, ntree = 1000)
rsf_fit_best <- result_rsf_train$best_model
rsf_fit_full <- rfsrc(Surv(t_dmfs, e_dmfs) ~ ., data = df, ntree = 1000)

rsf_v_importance <- result_rsf_train$importance  
kable_table_func(rsf_v_importance, "Table 12: Variable Importance from Random Survival Forest model")
```

# Model Discrimination Assessment Metrics

In this study, the discriminatory power of the survival model was assessed using the **average C-index** (Concordance Index) and **average iAUC** (Integrated Area Under the Curve) obtained through repeated cross-validation.

## 1. Concordance Index (C-index)

### Calculation Process

The C-index is calculated by randomly selecting a pair of individuals $(i, j)$. The pair is considered **concordant** if:

-   Individual $i$ has an actual event time shorter than individual $j$, **and**
-   The model assigns a higher risk score (or shorter predicted survival time) to individual $i$.

### Formula

The C-index represents the proportion of concordant pairs among all comparable pairs:

$$
C = \frac{\text{Number of concordant pairs}}{\text{Number of comparable pairs}}
$$

### Interpretation of Values

-   $C = 0.5$ → Prediction is no better than random guessing
-   $C = 1.0$ → Perfect discriminatory ability
-   $0.7 \leq C < 0.8$ → Acceptable discrimination
-   $0.8 \leq C < 0.9$ → Excellent discrimination
-   $C \geq 0.9$ → Outstanding discrimination

The C-index primarily assesses **risk ordering** but does not account for **time-dependent dynamics** in model performance.

## 2. Integrated AUC (iAUC)

The **iAUC** (Integrated Area Under the Curve) is the **integrated form of time-dependent AUC**. It measures a model's **discriminatory ability at different time points** and provides an **overall summary score** through integration (weighted averaging) over time.

The calculation proceeds in two main steps:

1.  **Compute time-dependent AUC(**$t$):

    -   At a specific time point $t$, define two groups:

        -   The **"event group"**: individuals who experienced the event by time $t$.
        -   The **"non-event/survival group"**: individuals who remain event-free beyond time $t$ (including those censored after $t$).

    -   Calculate the standard AUC (Area Under the ROC Curve) for classifying these two groups at time $t$ based on the model's predicted risk scores. Since the risk scores are fixed, at each time point $t$ we simply reclassify the samples into the event group and the survival group. We then compare the risk scores to see whether the Cox model correctly ranks those who experienced the event before $t$ ahead of those who survived beyond $t$.

2.  **Integrate over all observed time points**: The overall iAUC is obtained by integrating the time-dependent AUC($t$) function over the entire follow-up period $[0, \tau]$:

    $$
    \text{iAUC} = \int_{0}^{\tau} \text{AUC}(t) \cdot w(t)  \, dt
    $$

    Where:

    -   $\tau$ is the maximum observation time.
    -   $w(t)$ is a weight function, often based on the **distribution of event times** or the **size of the risk set** at time $t$ to account for the varying precision of the estimate across time.

### 

-   $\text{iAUC} = 0.5$ → Discriminatory power equivalent to **random guessing**.
-   $\text{iAUC} = 1.0$ → **Perfect discriminatory ability** across all time points.
-   Values between 0.5 and 1.0 indicate the degree of predictive accuracy, with higher values representing better performance.

## 3. Comparative Summary

| Feature | **C-index** | **iAUC** |
|:-----------------------|:-----------------------|:-----------------------|
| **Summary Type** | Global, single number | **Integrated** summary over time |
| **Perspective** | **Ranking** | **Prediction** at various time points |
| **Handling of Time** | Implicit, based on event order | Explicit, evaluates performance at each time $t$ |
| **Sensitivity to PH** | Less sensitive to minor violations of proportional hazards | Can be more informative if discrimination changes over time |
| **Calculation** | Based on **comparable pairs** | Based on **time-dependent ROC curves** |
| **Complexity** | Simpler to compute and interpret | More computationally intensive |

Both the C-index and iAUC provide valuable insights into model performance: - The **C-index** offers a straightforward measure of a model's overall ability to rank patients by risk - The **iAUC** provides a more comprehensive view of time-dependent predictive performance

The combination of these metrics through repeated cross-validation ensures a robust assessment of the survival model's discriminatory power.

```{r}
# prepare the test data

# evaluate the full Cox model 
result_valid <- calculate_time_auc_cindex(
        "Cox", 
        fitted_model = results_train$model, 
        df = clinical_cleaned_risk_test
      )

result_valid_cox_mc <- calculate_time_auc_cindex(
        "Cox", 
        fitted_model = results_train_cli$model, 
        df = clinical_cleaned_risk_test
      )
      
# evaluate the Random Survival Forest model
result_rsf_valid <- calculate_time_auc_cindex(
  "RSF",
  fitted_model = rsf_fit_full, #rsf_fit_best
  df = clinical_cleaned_risk_test
  )

# model comparision
comparison_df <- data.frame(
  Model = c("COX Clinical-Genomic", "RSF Clinical-Genomic", "COX Clinical-only"),
  iAUC = c(result_valid$iAUC, result_rsf_valid$iAUC, result_valid_cox_mc$iAUC),
  C_Index = c(result_valid$c_index, result_rsf_valid$c_index, result_valid_cox_mc$c_index),
  Sum_iAUC_Cindex = c(result_valid$iAUC, result_rsf_valid$iAUC, result_valid_cox_mc$iAUC) + c(result_valid$c_index, result_rsf_valid$c_index, result_valid_cox_mc$c_index)
)

kable(comparison_df, caption = "Table 13: Best Model Performance Comparison")

saveRDS(results_train$model, "best_cox_mm_model_1470.rds")
saveRDS(results_train_cli$model, "best_cox_mc_model_1515.rds")
# Because PH test for Cox doesn't hold, finally rsf_fit_full is taken as finial model.
saveRDS(rsf_fit_full, "best_rsf_model_mm_1489.rds")
```

To compare the performances of the Cox proportional hazards model and the Random Survival Forest model, Each algorithm will be trained and evaluated using three separate feature sets: an exclusive clinical variable set, an exclusive gene expression variable set, and a combined set integrating both clinical and genomic features. Through bootstrap cross validations with replacement, we calculate the everage of C_index and iAUC.

```{r}
geo_accession_7390 <- clinical_cleaned_7390$geo_accession
expr_mat_7390 <- expr[, geo_accession_7390, drop = FALSE]

# res7390_3_model4 <- run_bootstrap_validation_3_model(expr_mat = expr_mat_7390,
#                                                      clinical_df = clinical_cleaned_7390,
#                                                      B = 100, 
#                                                      seed = 92000000, 
#                                                      min_epv = 5, 
#                                                      coef_max = 10, 
#                                                      min_concord = 0.97)

# Create a comparison dataframe
res7390_3_model4 <- readRDS("res7390_3_model4.rds")

model_comparison_mean <- data.frame(
  Model_Type = c("Clinical-only", "Genomic-only", "Clinical-Genomic Integrated"),
  Mean_Cox_iAUC = c(
    res7390_3_model4$summary_results_mc$mean_cox_iAUC,
    res7390_3_model4$summary_results_mg$mean_cox_iAUC,
    res7390_3_model4$summary_results_mm$mean_cox_iAUC
  ),
  Mean_Cox_Cindex = c(
    res7390_3_model4$summary_results_mc$mean_cox_cidx,
    res7390_3_model4$summary_results_mg$mean_cox_cidx,
    res7390_3_model4$summary_results_mm$mean_cox_cidx
  ),
  Mean_RSF_iAUC = c(
    res7390_3_model4$summary_results_mc$mean_rsf_iAUC,
    res7390_3_model4$summary_results_mg$mean_rsf_iAUC,
    res7390_3_model4$summary_results_mm$mean_rsf_iAUC
  ),
  Mean_RSF_Cindex = c(
    res7390_3_model4$summary_results_mc$mean_rsf_cidx,
    res7390_3_model4$summary_results_mg$mean_rsf_cidx,
    res7390_3_model4$summary_results_mm$mean_rsf_cidx
  )
)

model_comparison_mean$Cox_Sum <- model_comparison_mean$Mean_Cox_iAUC + model_comparison_mean$Mean_Cox_Cindex
model_comparison_mean$RSF_Sum <- model_comparison_mean$Mean_RSF_iAUC + model_comparison_mean$Mean_RSF_Cindex

# Print the comparison table
kable(model_comparison_mean, caption = "Table 14: Model Performance Comparision (Mean) from 100 Bootstrap Iterations") 

model_summary_df <- res7390_3_model4$summary_df %>%
  rename(model_type = model)
model_summary_df$model_type <- c(
  "mc (Clinical-only)",
  "mg (Genomic-only)", 
  "mm (Clinical-Genomic Integrated)"
)

kable(model_summary_df, caption = "Table 15: Model Best Performance from 100 Bootstrap Iterations") 
```

Across 100 bootstrap resamples, the clinical-only Cox model exhibited a slight performance advantage over the alternative models; however, its mean iAUC and mean C-index were only 0.55 and 0.52, respectively, indicating limited discriminative ability. If one were to nominate a single model for prospective prediction, the clinical-only Cox model achieved the highest combined score in bootstrap replicate 28 (Cox_iAUC + Cox_Cindex = 1.52). Nevertheless, because the proportional hazards assumption did not hold for this model (PH test), the runner-up—the clinical–genomic integrated random survival forest (RSF)— is selected as the final model.

```{r}
best_rsf_model_mm_1489 <- readRDS("best_rsf_model_1489.rds")
```

```{r}

expr_df <- as.data.frame(t(expr))
#combine gene with clinical data
clinical_t_e <- clinical_cleaned_7390 %>% select(geo_accession, t_dmfs, e_dmfs)
merged_df <- expr_df %>%
  rownames_to_column("geo_accession") %>%
  left_join(clinical_t_e, 
            by = "geo_accession") %>%
  column_to_rownames("geo_accession")




# Set number of iterations
n_runs <- 100

# Create empty vectors to store performance metrics for each run
c_index_results <- numeric(n_runs)
cox_me_cindex_results <- numeric(n_runs)
iAUC_results <- numeric(n_runs)
total_score_results <- numeric(n_runs)

# Create a list to store all complete results (optional, if memory permits)
all_results_list <- list()

# Initialize variables to track the best results
best_total_score <- -Inf  # Initialize to negative infinity
best_run_id <- 0
best_results <- NULL

# Run 100 iterations
for (i in 1:n_runs) {
  # Set different random seeds to ensure different data splits each time
  current_seed <- 123456 + i
  set.seed(current_seed)
  
  cat(sprintf("Running iteration %d/%d...\n", i, n_runs))
  
  # Execute the model training function
  results <- train_cox_with_pca(
    data = merged_df, 
    time_col = "t_dmfs", 
    status_col = "e_dmfs", 
    clinical_cleaned_7390, 
    train_ratio = 0.7, 
    variance_threshold = 0.7
  )
  
  # Store performance metrics
  c_index_results[i] <- results$c_index
  cox_me_cindex_results[i] <- results$cox_me_cindex
  iAUC_results[i] <- results$iAUC
  total_score_results[i] <- results$total_score
  
  # Store complete results in list (optional)
  all_results_list[[i]] <- results
  
  # Check if this is the best result so far
  if (results$total_score > best_total_score) {
    best_total_score <- results$total_score
    best_run_id <- i
    best_results <- results  # Save the complete best result object
    
    cat(sprintf("  New best result found! Run ID: %d, Total Score: %.4f\n", i, best_total_score))
  }
  
  # Print progress
  cat(sprintf("  Completed: C-index = %.3f, iAUC = %.3f, Total = %.3f\n", 
              results$c_index, results$iAUC, results$total_score))
}



# Calculate mean and standard deviation of performance metrics
mean_c_index <- mean(c_index_results)
sd_c_index <- sd(c_index_results)
mean_coxme_cindex <- mean(cox_me_cindex_results)
sd_coxme_cindex <- sd(cox_me_cindex_results)
mean_iAUC <- mean(iAUC_results)
sd_iAUC <- sd(iAUC_results)
mean_total_score <- mean(total_score_results)
sd_total_score <- sd(total_score_results)


pca_avg_results <- data.frame(
  Metric = c("C-index", "CoxMe C-index", "iAUC", "Total Score"),
  Mean   = c(mean_c_index, mean_coxme_cindex, mean_iAUC, mean_total_score),
  SD     = c(sd_c_index, sd_coxme_cindex, sd_iAUC, sd_total_score)
)

kable(pca_avg_results, caption = "Table 16: PCA Cox Model Performance (Mean) from 100 Iterations") 

pca_best_results_df <- data.frame(
  Item   = c("Best Run ID", "Best Seed", "C-index", "iAUC", "Total Score",
             "Number of PCs", "Cumulative Variance (%)", "Global PH Test p-value"),
  Value  = c(best_run_id,
             12345 + best_run_id,
             best_results$c_index,
             best_results$iAUC,
             best_results$total_score,
             best_results$max_vars_allowed,
             best_results$cumulative_variance_explained * 100,
             best_results$global_ph_test)
)

kable(pca_best_results_df, caption = "Table 17: PCA Cox Model Best Performance ") 


saveRDS(best_results, "Best PCA_COX model.rds")

```

```{r}

model_comparison_mean
model_summary_df
#PCA
pca_best_results_df
pca_avg_results
```

standardization. Perform unsupervised clustering (e.g., K-means or hierarchical clustering) on gene expression data to identify potential molecular subgroups.

```{r}
#| echo: false
#| #| warning: false
#| 
# Load necessary libraries
#install.packages(c("pheatmap")) # Uncomment to install
library(survival)     # For survival analysis
library(survminer)    # For visualization of survival analysis
library(pheatmap)     # For creating heatmaps

# ------------------------------------------------------------------------------
# STEP 1: Data Preparation and Preprocessing
# ------------------------------------------------------------------------------

# Ensure the expression matrix and clinical data have matching sample IDs

different_samples <- setdiff(colnames(expr), clinical_cleaned_7390$geo_accession)
length(different_samples) == 0


# Feature selection: Identify most variable genes for clustering
# Using genes with highest variance across samples reduces noise
gene_vars <- apply(expr, 1, var) # Calculate variance for each gene (row)
top_gene_index <- order(gene_vars, decreasing = TRUE)[1:1500] # Select top 1500 most variable genes
expr_filtered <- expr[top_gene_index, ] # Subset expression matrix

# ------------------------------------------------------------------------------
# STEP 2: Hierarchical Clustering of Samples
# ------------------------------------------------------------------------------

# Calculate distance matrix between samples(patients) features are genes
# Note: dist() computes distance between rows, so we transpose the matrix
# to make samples (formerly columns) become rows
sample_dist <- dist(t(expr_filtered), method = "euclidean") 
# Alternative: Correlation-based distance (common in genomics)
# sample_dist <- as.dist(1 - cor(expr_filtered, method = "pearson"))

# Perform hierarchical clustering using Ward's method
# Ward's method minimizes within-cluster variance, producing compact clusters
sample_hclust <- hclust(sample_dist, method = "ward.D2")

# Visualize the dendrogram to determine appropriate number of clusters (K)
plot(sample_hclust, 
     labels = FALSE,          # Hide sample labels for clarity
     main = "Sample Hierarchical Clustering Dendrogram",
     xlab = "Samples", 
     ylab = "Height (Distance)")





# ------------------------------------------------------------------------------
# STEP 3: Cut Dendrogram to Define Molecular Subtypes
# ------------------------------------------------------------------------------

# Cut the dendrogram to obtain cluster assignments
# Choose k based on dendrogram structure and biological rationale

# Initialize variables to store results
p_values <- numeric(30)  # Store p-values for k=2 to 50
best_k <- NULL
best_p_value <- Inf

# Iterate over k from 2 to 50
for (k in 2:31) {
  # Cut dendrogram to define k molecular subtypes
  clusters <- cutree(sample_hclust, k = k)
  
  # Add cluster information to clinical data
  clinical_cleaned_7390$Molecular_Subtype <- factor(clusters, 
                                                   labels = paste0("Subtype_", 1:k))
  
  # Create survival object
  surv_obj <- Surv(time = clinical_cleaned_7390$t_dmfs, 
                  event = clinical_cleaned_7390$e_dmfs)
  
  # Fit survival curves
  fit <- survfit(surv_obj ~ Molecular_Subtype, data = clinical_cleaned_7390)
  
  # Perform log-rank test and extract p-value
  logrank_test <- survdiff(surv_obj ~ Molecular_Subtype, data = clinical_cleaned_7390)
  p_value <- 1 - pchisq(logrank_test$chisq, df = length(logrank_test$n) - 1)
  
  # Store p-value
  p_values[k-1] <- p_value
  
  # Update best k if current p-value is lower
  if (p_value < best_p_value) {
    best_p_value <- p_value
    best_k <- k
  }
  
  # Print progress
  cat(sprintf("k = %d: p-value = %.4f\n", k, p_value))
}

# Print final results
cat(sprintf("\nOptimal k: %d with minimum p-value: %.6f\n", best_k, best_p_value))

# Create results dataframe
results_df <- data.frame(
  k = 2:31,
  p_value = p_values
)

# Plot p-values vs k
library(ggplot2)
ggplot(results_df, aes(x = k, y = -log10(p_value))) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = best_k, linetype = "dashed", color = "red") +
  annotate("text", x = best_k, y = max(-log10(p_values)), 
           label = paste("Optimal k =", best_k), color = "red") +
  labs(title = "P-value vs Number of Clusters (k)",
       x = "Number of Clusters (k)",
       y = "-log10(p-value)") +
  theme_minimal()


k <- 5 # Example: define 2 molecular subtypes
clusters <- cutree(sample_hclust, k = k)

# Add cluster information to clinical data as a new factor variable
clinical_cleaned_7390$Molecular_Subtype <- factor(clusters,
                                              labels = paste0("Subtype_", 1:k))

surv_obj <- Surv(time = clinical_cleaned_7390$t_dmfs,
                 event = clinical_cleaned_7390$e_dmfs)

# Fit survival curves stratified by molecular subtype
fit <- survfit(surv_obj ~ Molecular_Subtype, data = clinical_cleaned_7390)

# Generate Kaplan-Meier survival plot with statistical analysis
surv_plot <- ggsurvplot(fit,
           data = clinical_cleaned_7390,
           pval = TRUE,
          # pval.coord = c(50, 0.9),
           conf.int = FALSE,
           risk.table = TRUE,
           risk.table.height = 0.25,
           legend.labs = levels(clinical_cleaned_7390$Molecular_Subtype),
           legend.title = "Molecular Subtype",
           palette = "jama",
           xlab = "Time (Days)", 
           ylab = "Overall Survival Probability",
           title = "Survival Analysis by Molecular Subtype",
           ggtheme = theme_minimal())

print(surv_plot)

```

Use t-SNE or UMAP for 2D visualization of clustering patterns to aid interpretation of subgroup structure.

```{r}


```

Feature Selection (if proportional hazards (PH) assumption holds) Method 1: Univariate Cox analysis to identify survival-associated genes Method 2: LASSO-Cox regression to select key predictive genes

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
